---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# VizSeasonalGams

<!-- badges: start -->
<!-- badges: end -->

The goal of VizSeasonalGams is to streamline plotting fitted gams that incorporate a seasonal component (e.g., smooths across day of year). While the tools here can work for models that include a smooth of interest across any continuous variable, this documentation will default to assuming the data and model include `"doy"` which contains day of year. You can provide an alternate variable name with the optional `plot_by` variable.

## Installation

You can install the development version of VizSeasonalGams from [GitHub](https://github.com/) with:

``` r
pak::pak("cbedwards-dfw/VizSeasonalGams")
```

## Example

Generate some simulated data akin to the kind of fish data this package was developed for, and fit a gam model to it:

```{r}
library(VizSeasonalGams)
library(mgcv)
library(tidyverse)

data.sim = simulate_data()

out = gam(fish_count ~ mark_stage +
            s(doy, by = mark_stage, k = 20) + 
            yearfac,
          method = "REML",
          family = "nb",
          data = data.sim)
```

`simulate_gam()` combines these two steps, but it seems valuable to have the model structure visible.

`VizSeasonalGams` package provides two key tools for working with fitted gam models that smooth across day of year. `plot_seasonal_gam_panels()` creates a tibble, where each row
is a unique combination of the levels of all factors represented in the model. The
`$plot` column contains a plot of the gam predictions for that combination of factors, setting any numeric predictaors aside from "doy" (or other specified `plot_across` variable) to their median. This allows easy exploration of the predictions of complex models over complex data.

`grid_panels()` (and to a lesser extent `wrap_panels()`) take the output of `plot_seasonal_gam_panels()` OR a subset thereof (e.g., filtering to levels of interest) and plots them as a grid or as panel. Think of them as analogs to facet_grid() and facet_wrap(), but for working with gam predictions generated by `plot_seasonal_gam_panels()`. 

### Create a dataframe of predictors and prediction panels
```{r}
prediction_panels = plot_seasonal_gam_panels(out)
prediction_panels
```

### Plot all prediction panels

We can simply wrap the panels:
```{r}
wrap_panels(prediction_panels)
```

This suits filtering our predictions to some cases of interest:

```{r}
prediction_panels |> 
  filter(yearfac == "2021") |> 
  wrap_panels()
```

If we have at least two factor predictors, we can plot a grid of these panels. Make sure to specify the variable names that define rows and columns!

```{r}
prediction_panels |> 
  grid_panels(column_var = "yearfac", row_var = "mark_stage")
```

We can also provide a "color by" variable to `plot_seasonal_gam_panels`; this will put additional information on each panel and reduce the number of panels. Because this reduces the dimensionality of `plot_seasonal_gam_panels()` output, it may make it impossible to use `grid_panels()` (ie only 1 factor predictor to plot by). `wrap_panels()` is a flexible solution.

```{r}
plot_seasonal_gam_panels(out, color_by = "yearfac") |> 
  wrap_panels()
```



